\Large \bf Web-based Data Analytics in Haskell

### 2011-12-17 paper/paper.tex
Added presentation HTML file and made more paper changes. Paper almost done now

\begin{document}
\vspace{-5cm}
\title{\Large \bf Web-based Data Analytics in Haskell}

@@

\begin{abstract}
Many people are interested in statistics, but providing an easy to use Website to analyze statistics can be difficult. We wanted to make it simple to create Websites to analyze arbitrary data. Our project explored data analytics using the Haskell Web framework, Yesod. We wrote a proof-of-concept site that analyzes information from the popular on-line game League of Legends. The Website allows users to lookup a variety of statistics and perform queries on the data while presenting the results in a graphical format.
\end{abstract}

@@

Analyzing statistics can be rewarding and insightful, but without an easy way to analyze new datasets doing so can be difficult. Statistics are useful in a verity of fields. For example, although no knowledge of statistics is required to play baseball, many fans of the sport love to analyze them. Since baseball is popular there are many Websites to analyze baseball data, but what about less popular data? Creating a site to analyze new sets of data can be time consuming. Although there are many existing statistics packages, they are often hard to use and do not provide a Website for others to browse. Users are often left managing large spreadsheets to keep track of their data.

% While the data being analyzed across different datasets may differ considerably, the sorts of queries that people are interested in are often similar. People would like to correlate different variables, filter on certain conditions, and retrieve subsets of the data. They would like to do calculations on the data (sums, averages, counts) and display this information in tables or charts.

We created a demo application using data from the on-line video game, League of Legends~\cite{lol}. League of Legends is a competitive Real Time Strategy games where teams of players compete against each other. The site allows users to view player information and analyze various gameplay statistics. Users can look at their own data to learn about their behavior in the game or look at more general data to see trends in gameplay. To create the site, we also had to write software to parse and upload League of Legends data.

Our aim in creating this application was to determine what sort of database interface would help us write this Website. We used the MongoDB\cite{mongo} database as our backend, and wrote code to automatically generate MongoDB queries based on columns defined on our dataset. Our paper discusses the issues, limitations, and advantages of our approach.

The rest of the paper is organized as follows. Section \ref{sec:background} provides some background. We discuss our package in section \ref{package}. Section \ref{site} looks at our example site. In section \ref{future} we discuss possible future directions of our work. Finally, section \ref{conclusion} concludes.

@@

Yesod~\cite{yesod} is one of several Web frameworks written in Haskell. Features of Yesod include templates, a persistence model, etc. Writing a Website in a Haskell framework allows developers to take advantage of Haskell features, such as type safety, which helps them write secure sites.

Yesod uses both MVC and REST. MVC is a paradigm for writing Websites that separates data and data logic (model) from the presentation of the data (view). REST is a convention for URLs so that sites can have a consistent pattern for accessing resources.

To write controllers, developers can use Yesod's monads, Handler and Widget. These build up HTML, CSS, and Javascript, and allow modifying and reading HTTP headers. These monads are instance of MonadIO which allows controller pages to execute arbitrary IO actions as well as generating the page. See Figure \ref{controller} for an example controller

Like most other frameworks, Yesod provides templates to make writing HTML, CSS, and JavaScript easier. The templates provide basic variable substitution and control statements, as well as support for simplified stuff. See Figure \ref{hamlet} for an example of Hamlet, Yesod's HTML templating system, which is used in the controller in Figure \ref{controller}.

Yesod uses Haskell's types to provide functionality that often are impossible in other languages. Type safety prevents many security vulnerabilities by ensuring that malicious code isn't used. Type inference allows more concise code by automatically detecting what model the developer is querying.

\begin{figure}[]
\footnotesize{
\begin{verbatim}
getGameIndexR :: Handler RepHtml
getGameIndexR = do
  (games, opts) <- paginateSelectList 10 [] []
  defaultLayout $ do
    let list = $(widgetFile "game/list")
    setTitle "Game Index"
    toWidget [cassius|
      ##header {
        float: left;
        color = #{color}}
    |]
    $(widgetFile "game/index")
\end{verbatim}
}
    \caption{Sample Yesod Controller}
    \label{controller}
\end{figure}

@@

Persistent is a library developed for Yesod to facilitate data access.  It attempts to make data access type safe, concise, declarative and independent of the underlying data store.  It uses a mixture of typeclasses to achieve this, as well as various language extensions to make some of tricks possible.  The three major pieces of the puzzle are the type classes \code{PersistBackend b m}, \code{PersistField val}, and \code{PersistEntity val}.

@@

\begin{figure}[]
\footnotesize{
\begin{verbatim}
data EntityField (GameGeneric backend) typ
    = typ ~ (Key backend (GameGeneric backend)) => GameId
    | typ ~ UTCTime   => GameCreated
    | typ ~ Bool      => GameRanked
    | typ ~ Int       => GameGameId
    | typ ~ Int       => GameLength
    | typ ~ GameStats => GameGameStats
    | typ ~ Text      => GameTeamPlayerSummoner Text
    | typ ~ Text      => GameOtherTeamPlayerSummoner Text
\end{verbatim}
}
    \caption{EntityField declarations for GameGeneric backend}
    \label{EntityField}
\end{figure}

@@

Filters are created through comparison functions, which are analogs for familiar comparison functions, such as less-than, equals and set membership.  They are defined by the \code{PersistFilter} data type, and Yesod provides more convenient creation functions such as \code{<.} and \code{==.}.  The filters also ensure type-safety of the comparison (Figure \ref{Filter}).  Note the Rank2Type for \code{typ} which enforces a match between the value and the field's type.  Also note that \code{v} represents the model type, which ensures all filters correspond to columns on the same data type.
\begin{figure}[]
\footnotesize{
\begin{verbatim}
data Filter v = forall typ . PersistField typ =>
    Filter { filterField :: EntityField v typ
           , filterValue :: Either typ [typ]
           , filterFilter :: PersistFilter
           }
    | FilterAnd [Filter v]
    | FilterOr [Filter v]
\end{verbatim}
}
    \caption{EntityField declarations for GameGeneric backend}
    \label{Filter}
\end{figure}
Select queries naturally follow from their filters, which encode almost all information needed to make the query.  The select only requires a few additional options such as limits and offsets before it can be executed.

@@

\subsection{Flot}

To generate charts, we use Flot~\cite{flot}, a JavaScript charting library. We originally used the Chart package, but switched to Flot for compatibility reasons and because JavaScript charts are more dynamic. Since Flot did not have Haskell bindings, we had to write our own.

@@

\section{Our project}
\label{package}

@@

\begin{figure*}[t]
\footnotesize{
\begin{verbatim}
type Game = GameGeneric backend
type GameId = Key backend Game
GameCreated :: EntityField Game UTCTime
GameGameStats :: EntityField Game GameStats

selectList :: (PersistEntity val, PersistBackend b m) => [Filter val] -> [SelectOpt val]
              -> b m [(Key b val, val)]
selectList [GameCreated <. time] [LimitTo 10] :: PersistBackend b m => b m [(GameId, Game)]
\end{verbatim}
}
    \caption{Query stuff}
    \label{querycode}
\end{figure*}

@@

\begin{figure*}[t]
\footnotesize{
\begin{verbatim}
class PersistEntity model => Queryable model where
    -- Column definitions
    data QueryColumn model :: * -> *
    queryColumnName     :: QueryColumn model typ -> UString
    queryFilter         :: QueryColumn model typ -> Value -> Document
    queryKeyCode        :: QueryColumn model typ -> Javascript
    queryMapCode        :: QueryColumn model typ -> Javascript
    queryReduceCode     :: QueryColumn model typ -> Javascript
    queryFinalizeCode   :: QueryColumn model typ -> Javascript
    queryCastResult     :: (Val typ) => QueryColumn model typ -> Value -> Maybe typ
    queryMapFunc        :: QueryColumn model typ -> forall typ0. [QueryColumn model typ0] -> Javascript
    queryReduceFunc     :: forall typ0. [QueryColumn model typ0] -> Javascript
    queryFinalizeFunc   :: forall typ0. [QueryColumn model typ0] -> Javascript
    queryCollection     :: QueryColumn model typ -> UString
\end{verbatim}
}
    \caption{Map-Reduce API}
    \label{queryable}
\end{figure*}

@@

Our Website displays appropriate information using JavaScript charts. While some data is best presented in tabular form, charts can visually display the relationship of different variables in a way that is easy for people to understand. In particular, we thought viewing a chart would help users view correlations between two different variables. Figure \ref{chart} shows a sample chart.

We used the Flot JavaScript charting package. Flot charts are generated using JavaScript and displayed on an HTML Canvas element. While we had originally tried generating static charts, that did not give us the same level of functionality as the JavaScript charts. Although we didn't end up using any of this functionality, it would have allowed users to sort charts in the browser, click on different elements to go to the appropriate page, etc.

While it is possible for developers to manually generate JavaScript to create Flot charts, we created a Haskell API that allows developers to specify their charts in Haskell. The API takes a list of display parameters (such as titles, chart size, orientation, etc.) and a list of data points in the form \code{[(key, value)]} .The chart automatically takes care of including any necessary HTML, JavaScript, and CSS and can simply be included anywhere in the page. Figure \ref{chartcode} shows a page that uses the chart API. \code{myChart} sets a few chart options and the data is queried from the database.

\begin{figure}[]
\footnotesize{
\begin{verbatim}
getFooR :: Handler RepHtml
getFooR = do
    data <- runDB blah buildQuery blah
        (QuerySomething foo) []
        [Query Something])
    let chart = makeChart myChart data
    ...

myChart :: Widget
myChart =
    setTitles "XAxis" "YAxis" "Title" $
    setHorizontal True $
    setSize 600 400 $
    barChart
\end{verbatim}
}
    \caption{Chart API}
    \label{chartcode}
\end{figure}

@@

@@

\begin{figure}[btp]
\footnotesize{
\begin{verbatim}
body = (com.riotgames.gameclient::EndOfGameStats)#1
  basePoints = 69
  completionBonusPoints = 0
  difficulty = (null)
  elo = 0
  eloChange = 0
  experienceEarned = 0
  gameId = 245368572
  gameLength = 1839
  gameMode = "CLASSIC"
  gameType = "MATCHED_GAME"
  newSpells = (mx.collections::ArrayCollection)#2
    filterFunction = (null)
    length = 0
    list = (mx.collections::ArrayList)#3
      length = 0
      source = (Array)#4
    sort = (null)
    source = (Array)#4
\end{verbatim}
}
    \caption{Sample ActionScript object}
    \label{samplelog}
\end{figure}

@@

The first phase of our parsing is to parse object dumps into an \code{ASValue} datatype, which is modeled after \code{Data.Aeson.Value} and described in figure \ref{asvalue}.  The parsing code is written with Attoparsec, and is straightforward other than the indentation.  Indentation is tracked by wrapping the \code{Parse} monad in a reader transformer that stores the number of spaces on the current indentation.  Every key-value pair of an object or entry in an array first checks for the corresponding level of indentation.  When a new object or array is encountered, the indentation is incremented for the recursive call, and if the indentation fails to parse it simply signifies the end of the object and the \code{many} combinator will return the results parsed so far.

@@

To test our data analytics package, we wrote a Website to display League of Legends statistics. Although the Website is fairly simple for an analysis Website, building the site helped us in developing our query interface. The Website allows users to upload games, view information about games, viwe champion data, and perform analyzis on summoner data.

Our summoner page, seen in \ref{todo}, displays aggregate information about a summoner across all uploaded games the summoner played in. The data is viewable in both table and chart format, and the columns are customizable about the user.

Our backend is capable of performing more powerful queries, but we did not have enough time to code more pages to display this functionality. Ideally, users would be able to specify arbitrary queries, giving custom filters, keys, etc.

In Figure \ref{list}, we can see a list of games. Probably want a more interesting picture though.

\begin{figure}[h]
    \includegraphics[width=80mm]{imgs/stats.png}
    \caption{Statistics for a summoner}
    \label{chart}
\end{figure}

>>>>

\begin{document}

\title{\Large \bf Web-based Data Analytics in Haskell}

@@

\begin{abstract}
Many people are interested in statistics, but providing an easy to use Website to analyze statistics can be difficult. We wanted to make it simple to create Websites to analyze arbitrary data. Our project explored data analytics using the Haskell Web framework, Yesod, using it to write a proof-of-concept site that analyzes information from the popular on-line game League of Legends. The Website allows users to lookup a variety of statistics and perform queries on the data while presenting the results in a graphical format.
\end{abstract}

@@

Analyzing statistics can be rewarding and insightful, but without an easy way to analyze new datasets doing so can be difficult. Statistics are useful in a variety of fields. For example, although no knowledge of statistics is required to play baseball, many fans of the sport love to analyze baseball statistics. Since baseball is popular there are many Websites to analyze baseball data, but what about less popular data? Creating a site to analyze new sets of data can be time consuming. Although there are many existing statistics packages, they are often hard to use and do not provide a Website for others to browse. Users are often left managing large spreadsheets to keep track of their data.

 While the data being analyzed across different datasets may differ considerably, the sorts of queries that people are interested in are often similar. People would like to correlate different variables, filter on certain conditions, and retrieve subsets of the data. They would like to do calculations on the data (sums, averages, counts) and display this information in tables or charts. We wanted to see if we could provide this functionality in a generic way for generic datasets.

We created a demo application using data from the on-line video game, League of Legends~\cite{lol}. League of Legends is a competitive Real Time Strategy games where teams of five play against each other. The site allows users to view player information and analyze various gameplay statistics. Users can look at their own data to learn about their behavior in the game or look at more general data to see trends in gameplay. Besides creating the site itself, we also wrote software to parse and upload League of Legends data.

Our aim in creating this application was to determine what sort of database interface would help us write this Website, as well as to evaluate Haskell's viability for Web application development. We chose to use the Yesod framework, which seemed to be the most mature Haskell Web framework. For our backend, we used the MongoDB\cite{mongo} database, and wrote code to automatically generate MongoDB queries based on columns defined on our dataset. Our paper discusses the issues, limitations, and advantages of our approach.

The rest of the paper is organized as follows. Section \ref{sec:background} provides some background on technologies we depended on for our project. We discuss our results in section \ref{results}. Section \ref{site} looks at our example site. In section \ref{future} we discuss possible future directions of our work. Finally, section \ref{conclusion} concludes.

@@

Yesod~\cite{yesod} is one of several Web frameworks written in Haskell. Like many other frameworks, Yesod uses both MVC and RESTful URLs, making it in some ways similar to more traditional frameworks. Being written in Haskell, however, Yesod allows developers to take advantage of unique Haskell features.

To write controllers, developers can use Yesod's monads, \code{Handler} and \code{Widget}. These build up HTML, CSS, and Javascript, and allow modifying and reading HTTP headers. These monads are instances of MonadIO which allows controller pages to execute arbitrary IO actions as well as generating the page. See Figure \ref{controller} for an example controller. This example queries a lists of games and paginates them into groups of 10. It then renders the page and includes a list template which shows information about each game. Part of this page can be seen in Figure \ref{list}.

Like most other frameworks, Yesod provides templates to make writing HTML, CSS, and JavaScript easier. The templates provide basic variable substitution and control statements, as well as supports for simplified syntax. See Figure \ref{hamlet} for an example of Hamlet, Yesod's HTML templating system, which is used in the controller in Figure \ref{controller}. The template resembles HTML but clearly is not HTML. Closing tags are not specified, Haskell functions are called, other templates are included, and static routes are used to generate URLs.

Yesod uses Haskell's types to provide functionality that is often impossible in other languages. Type safety helps prevent many security vulnerabilities and broken URLs. For example, since URLs are represnted by a special type in Yesod, it is impossible to refer to a URL which does not exist because this would be a type violation. Type inference allows more concise code by automatically detecting what model the developer is querying, as in Figure \ref{controller}. Yesod automatically detects that we want a list of games because how the \code{games} variable is later used.

\begin{figure}[]
\footnotesize{
\begin{verbatim}
getGameIndexR :: Handler RepHtml
getGameIndexR = do
  (games, opts) <- paginateSelectList 10 [] []
  defaultLayout $ do
    let list = $(widgetFile "game/list")
    setTitle "Game Index"
    $(widgetFile "game/index")
\end{verbatim}
}
    \caption{Sample Yesod Controller}
    \label{controller}
\end{figure}

@@

Persistent\cite{persistent} is a library developed for Yesod to facilitate data access.  It attempts to make data access type safe, concise, declarative and independent of the underlying data store.  It uses a mixture of typeclasses to achieve this, as well as various language extensions to make some of tricks possible.  The three major pieces of the puzzle are the type classes \code{PersistBackend b m}, \code{PersistField val}, and \code{PersistEntity val}.

@@

\begin{figure}[]
\scriptsize{
\begin{verbatim}
data EntityField (GameGeneric backend) typ
   = typ ~ (Key backend (GameGeneric backend)) => GameId
   | typ ~ UTCTime   => GameCreated
   | typ ~ Bool      => GameRanked
   | typ ~ Int       => GameGameId
   | typ ~ Int       => GameLength
   | typ ~ GameStats => GameGameStats
   | typ ~ Text      => GameTeamPlayerSummoner Text
   | typ ~ Text      => GameOtherTeamPlayerSummoner Text
\end{verbatim}
}
    \caption{EntityField declarations for GameGeneric backend}
    \label{EntityField}
\end{figure}

@@

Filters are created through comparison functions, which are analogs for familiar comparison functions, such as less-than, equals and set membership.  They are defined by the \code{PersistFilter} data type, and Yesod provides more convenient creation functions such as \code{<.} and \code{==.}.  The filters also ensure type-safety of the comparison (Figure \ref{Filter}).  Note the Rank2Type for \code{typ} which enforces a match between the value and the field's type.  Also note that \code{v} represents the model type, which ensures all filters correspond to columns on the same data type.
\begin{figure}[]
\footnotesize{
\begin{verbatim}
data Filter v = forall typ . PersistField typ =>
    Filter { filterField :: EntityField v typ
           , filterValue :: Either typ [typ]
           , filterFilter :: PersistFilter
           }
    | FilterAnd [Filter v]
    | FilterOr [Filter v]
\end{verbatim}
}
    \caption{Filter typeclass}
    \label{Filter}
\end{figure}
Select queries naturally follow from their filters, which encode almost all information needed to make the query.  The select only requires a few additional options such as limits and offsets before it can be executed.

@@

@@

\section{Results}
\label{results}

@@

@@

\begin{figure*}[t]
\footnotesize{
\begin{verbatim}
class PersistEntity model => Queryable model where
    -- Column definitions
    data QueryColumn model :: * -> *
    queryColumnName     :: QueryColumn model typ -> UString
    queryFilter         :: QueryColumn model typ -> Value -> Document
    queryKeyCode        :: QueryColumn model typ -> Javascript
    queryMapCode        :: QueryColumn model typ -> Javascript
    queryReduceCode     :: QueryColumn model typ -> Javascript
    queryFinalizeCode   :: QueryColumn model typ -> Javascript
    queryCastResult     :: (Val typ) => QueryColumn model typ -> Value -> Maybe typ
    queryMapFunc        :: QueryColumn model typ -> forall typ0. [QueryColumn model typ0] -> Javascript
    queryReduceFunc     :: forall typ0. [QueryColumn model typ0] -> Javascript
    queryFinalizeFunc   :: forall typ0. [QueryColumn model typ0] -> Javascript
    queryCollection     :: QueryColumn model typ -> UString
\end{verbatim}
}
    \caption{Queryable Typeclass}
    \label{queryable}
\end{figure*}

@@

To help users understand the data they are looking at, our Website renders data using graphical charts, such as bar or line charts. While some data is best presented in tabular form, charts can visually display the relationship of different variables in a way that is easy for people to understand. In particular, we thought viewing a chart would help users view correlations between two different variables. Figure \ref{chart} shows a sample chart.

We used the Flot\cite{flot} JavaScript charting package. Flot charts are generated using JavaScript and displayed on an HTML Canvas element. While we had originally tried generating static charts, that did not give us the same level of functionality as the JavaScript charts. Although we didn't end up using any of this functionality, it would have allowed users to sort charts in the browser, click on different elements of a chart to go to the appropriate page, etc.

While it is possible for developers to manually generate JavaScript to create Flot charts, we created a Haskell API that allows developers to specify their charts in Haskell. The API takes a list of display parameters (such as titles, chart size, orientation, etc.) and a list of data points in the form \code{[(key, value)]}. The chart automatically takes care of including any necessary HTML, JavaScript, and CSS and can simply be included anywhere in the page, using Yesod's Widget monad. Figure \ref{chartcode} shows a controller that uses the chart API.

@@

\begin{figure}[]
\footnotesize{
\begin{verbatim}
summonerChartR :: Handler RepHtml
sumonnerChartR = do
    dataRows <- runDB . runMapReduce $
             buildQuery(QGameChampion summonerName)
        [exists $ QGameSummoner summonerName] columns
    let series = getSeries dataRows

    defaultLayout $ do
        setTitle $ "Stats for " ++ summonerName
        let stats = makeChart summonerChart series
        $(widgetFile "summoner/view")

summonerChart :: Widget
summonerChart =
    setTitles "" "" "Summoner Statistics" $
    setHorizontal True $
    setSize 600 400 $
    barChart
\end{verbatim}
}
    \caption{Controller used to generate a chart}
    \label{chartcode}
\end{figure}

@@

\begin{figure}[btp]
\footnotesize{
\begin{verbatim}
body = (com.riotgames.client::EndOfGameStats)#1
  basePoints = 69
  completionBonusPoints = 0
  difficulty = (null)
  elo = 0
  eloChange = 0
  experienceEarned = 0
  gameId = 245368572
  gameLength = 1839
  gameMode = "CLASSIC"
  gameType = "MATCHED_GAME"
  newSpells = (mx.collections::ArrayCollection)#2
    filterFunction = (null)
    length = 0
    list = (mx.collections::ArrayList)#3
      length = 0
      source = (Array)#4
    sort = (null)
    source = (Array)#4
\end{verbatim}
}
    \caption{Sample ActionScript object}
    \label{samplelog}
\end{figure}

@@

The first phase of our parsing is to parse object dumps into an \code{ASValue} datatype, which is modeled after \code{Data.Aeson.Value} and described in Figure \ref{asvalue}.  The parsing code is written with Attoparsec, and is straightforward other than the indentation.  Indentation is tracked by wrapping the \code{Parse} monad in a reader transformer that stores the number of spaces on the current indentation.  Every key-value pair of an object or entry in an array first checks for the corresponding level of indentation.  When a new object or array is encountered, the indentation is incremented for the recursive call, and if the indentation fails to parse it simply signifies the end of the object and the \code{many} combinator will return the results parsed so far.

@@

We wrote a Website to display League of Legends statistics to determine how well our Peristence integration worked. Although the Website is fairly simple for an analysis site, building the site helped us in developing our query interface. The Website allows users to upload games, view information about games, viwe champion data, and perform analyzis on summoner data. When a game is uploaded, its data is added to the MongoDB database automatically and will be included in all future queries.

Figure \ref{list} shows a list of games, indicating which Champions were used in each match. Figure \ref{chart} shows a sample chart that shows aggregate information about a summoner across all uploaded games the summoner played in, while Figure \ref{stats} shows the same information but in tabular format. The presentation of the results as well as the columns of interest are customizable by the user. In writing these pages, we found that we had to write a lot of code to transform data to and from a variety of formats, but this was fairly easy using Haskell's functional list manipulation.

Our backend is capable of performing more powerful queries, but we did not have enough time to code more pages to display this functionality. Ideally, users would be able to specify arbitrary queries, giving custom filters, keys, etc. Questions we'd like users to be able to answer include ``How does X correlate to Y for players above a certain skill level?'', ``What is the distribution of game length for different champions?'', or ``Rank all players by the quality of their wins and losses''.

\begin{figure}[h]
    \includegraphics[width=80mm]{imgs/stats.png}
    \caption{Statistics for a summoner}
    \label{stats}
\end{figure}
